### 2016-11-14 Paper_v1.tex
Nov14
Nov14_2

The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (pyruvate being consumed to produce lactate) (Fig.~\ref{fig:Sensitivity}, bottom, section F).
The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation, and 1 pyrophosphatase.
The system state has even more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
For example, the first-order effect of alanine synthesis is very high; it consumes both pruvate and glutamate, two key species in the network.
In addition, a handful of alanine synthesis pairwise sensitivities are high.
However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model.
Malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
Thus, the total-order alanine synthesis sensitivity is low, placing it at the very bottom of section I.
Another interesting result of this analysis is seen in the intersection of sections F and G with section J.
The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0); therefore, the perturbation of these reactions (multiplying the rate constant by 1.01) had no actual effect on the model.
Thus, all pairwise sensitivities with reactions in section J can essentially be considered first-order sensitivities for the other reactions.
Interestingly, the reactions in section F and several in section G showed most of thier highest sensitivities when paired with the "non-effects" of section J.
Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO4 transport.
This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.

\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
%Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
However, we first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig. ~\ref{fig:CATProdYield}A). 
The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig. ~\ref{fig:CATProdYield}B).
The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
Whereas, the case constrained by experimental data had a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
In comparing the flux distributions between the unconstrained and constrained cases (Fig. ~\ref{fig:Network}), the constrained cases heavily utilize the first step in the pentose phosphate pathway to generate NADPH.
The majority of the flux continues through the Entner–Doudoroff pathway whereas in the unconstrained case, the majority of flux travels through glycolysis.
In all cases, the energy source comes from oxidative phosphorylation with some from the citric acid cycle.
Whereas, in just the TX/TL case, there is a high flux through fumerate dehydrogenase from aspartic acid uptake. 
In the unconstrained and most constrained case, we see a mixture of acetate and lactate accumulation.
This shows the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
As a result, high oxidative phosphorylation activity relative to our cell free system leads to an acetate overflow.
This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB} which is responsible for regenerating NADPH.
The simulation showed carbon was diverted towards lactate, however since CAT production is constrained by the translation rate, we expected there to be no increase in CAT production.
The decrease in acetate production is a promising result to increase CAT yield. 
A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase, showed carbon being diverted towards lactate and succinate, however it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.

@@

In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance. 
The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case. 
Knockouts of \textit{gnd} and phosphate acetyltransferase, show carbon can be diverted away from acetate and potentially towards CAT or other proteins of interest expressed in CFPS. 
The other limitations to address in CFPS would be to enhance the transcription and translation rates since the protein of interest to be expressed is ultimately bounded by these kinetic rates.    
Li et al. have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA and tRNAs \cite{2014_li_PlosOne}. 
Underwood et al. has also shown the increase in ribosome levels does not significantly increase protein yields and rates, however adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
In addition to improving CFPS performance, Jewett et al. has showed that oxidative phosphorylation operates in cell-free systems and knocking out oxidative phosphorylation reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}. 
However, it is inconclusive how much oxidative phosphorylation activity there is compared to \textit{in vivo} systems and both of our models suggest oxidative phosphorylation is vital to CAT production. 
Thus, this is a potential place for improvement to optimize CFPS for better performance and protein yield.

@@

\subsection*{Local sensitivity analysis}
We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=norm(CAT(p_i,p_j,t)-CAT(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of CAT production to the $i$th and $j$th parameters, $CAT(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i$th and $j$th parameters, $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.
The array of sensitivity indices was organized into a clustergram (Fig.~\ref{fig:Sensitivity}, top) with both axes corresponding to the 163 rate constants rearranged from most sensitive (top, right) to least sensitive (bottom, left).

Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).

@@

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.8\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
\caption{Normalized first-order and pairwise sensitivities of CAT production (top) and system state (bottom) to rate constants.}
\label{fig:Sensitivity}
\end{figure}

>>>>

The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (pyruvate being consumed to produce lactate) (Fig.~\ref{fig:Sensitivity}, bottom, section F).
The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation; and 1 pyrophosphatase.
The system state has even more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
For example, the first-order effect of alanine synthesis is very high; it consumes both pruvate and glutamate, two key species in the network.
In addition, a handful of alanine synthesis pairwise sensitivities are high.
However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model; malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
Thus, the total-order alanine synthesis sensitivity is low, placing it at the very bottom of section I.
Another interesting result of this analysis is seen in the intersection of sections F and G with section J.
The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0); therefore, the perturbation of these reactions (multiplying the rate constant by 1.01) had no actual effect on the model.
Thus, all pairwise sensitivities with reactions in section J can essentially be considered first-order sensitivities for the other reactions.
Interestingly, the reactions in section F and several in section G showed most of thier highest sensitivities when paired with the "non-effects" of section J.
Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO$_4$ transport.
This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.

\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
%Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
We first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig.~\ref{fig:CATProdYield}A). 
The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig.~\ref{fig:CATProdYield}B).
The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
Thus, the experimental dataset and best-fit parameter set each produced CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
However, the case constrained by experimental data showed a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
In comparing the flux distributions between the unconstrained and constrained cases (Fig.~\ref{fig:Network}), the constrained cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH.
In these cases the majority of the flux continued through the Entner–Doudoroff pathway, whereas in the unconstrained case the majority of flux traveled through glycolysis.
In all cases, the energy source came primarily from oxidative phosphorylation, as well as partly from the TCA cycle.
In the TX/TL constrained case, there was a high flux through fumerate dehydrogenase from aspartic acid uptake, whereas in the unconstrained and most constrained cases, acetate and lactate accumulation occurred.
This shows that the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
As a result, high oxidative phosphorylation activity relative to our cell-free system leads to an acetate overflow.
This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB}, which is responsible for regenerating NADPH.
The simulation showed carbon was diverted toward lactate; however, since CAT production is constrained by the translation rate, we expected no increase in CAT production.
The decrease in acetate production is promising as a mechanism to increase CAT yield. 
A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase showed carbon being diverted toward lactate and succinate; however, it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.

@@

In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance. 
The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case. 
Knockouts of \textit{gnd} and phosphate acetyltransferase, show carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS. 
The other limitations to address in CFPS would be to enhance the transcription and translation rates since the protein of interest to be expressed is ultimately bounded by these kinetic rates.    
Li et al. have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA and tRNAs \cite{2014_li_PlosOne}. 
Underwood et al. has also shown the increase in ribosome levels does not significantly increase protein yields and rates; however, adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
In addition to improving CFPS performance, Jewett et al. has showed that oxidative phosphorylation operates in cell-free systems and knocking out oxidative phosphorylation reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}. 
However, it is inconclusive how much oxidative phosphorylation activity there is compared to \textit{in vivo} systems and both of our models suggest oxidative phosphorylation is vital to CAT production. 
Thus, this is a potential place for improvement to optimize CFPS for better performance and protein yield.

@@

\subsection*{Local sensitivity analysis}
We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:CAT-sensitivity}
\begin{split}
    S_{ij}=norm(CAT(p_i,p_j,t)-CAT(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of CAT production to the $i$th and $j$th parameters, $CAT(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i$th and $j$th parameters, $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.
The array of sensitivity indices was organized into a clustergram (Fig.~\ref{fig:Sensitivity}, top) with both axes corresponding to the 163 rate constants rearranged from most sensitive (top, right) to least sensitive (bottom, left).

Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:system-sensitivity}
\begin{split}
    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth_png.pdf}
\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
\label{fig:CarbonBoth_png}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.76\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
\caption{Normalized first-order and pairwise sensitivities of CAT production (top) and system state (bottom) to rate constants.}
\label{fig:Sensitivity}
\end{figure}
